A Global Picture of Hunger

### 2020-11-17 docs/modeling_fies.tex
almost done with draft; improve validation

When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.  We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find reasonable accuracy ($R^2 = 0.799$, MAE = 0.068 for moderate food insecurity and $R^2 = 0.772$, MAE = 0.038 for severe food insecurity).  This suggestions that our model explains between 77\% to 79\% of the variation in food security outcomes and is on average, within 3-7 percentage points in its estimates of rates of food insecurity, when applied to countries outside of the training set.

@@

\subsection{Modeling}
We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.  We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.  A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees. Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.

To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.  Key hyperparameters include the number of decision trees to generate, the number of variables randomly selected as candidates for splitting a node, and the average number of observations in a leaf node.  We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate and severe models. By choosing the combinations of hyperparameters with the smallest OOB error we found the optimal combination for both models. We found that 5000 trees are more than sufficient for the error rates to converge.

After fitting the models, we examined the importance of the individual covariates in explaining rates of moderate and severe food insecurity. We used a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable}.  Similarly to model predictions, we transformed the error term with the inverse logit so that the errors would be on the scale of the outcome variable and therefore more interpretable.

\subsection{Model Validation}
We observed a MAE of 0.0029 for our model of severe food insecurity ($R^2$ of 0.9978) and an MAE 0.0054 for our model of moderate food insecurity ($R^2$ of 0.9981).  This means that, on average, our model predictions were only of by 0.29\% and 0.54\%, respectively, for the subnational areas in countries where we had microdata.

To further validate the models, we ran a ten-fold cross validation at the country level.  We randomly divided the countries into ten parts, and re-fit the model 10 times, with each part left out one.  Then, we predicted the values for the ten percent that was left out using the model fit with 90\% of the countries.  We randomly left out data at the country level because within each country we had values from different subnational areas and different years, and these values are likely to be very similar to each other.  Because we are extending our models predictions to countries that were not found at all in our training data, a better test of our model is to assess its predictive accuracy on countries that were entirely absent from the training data.  By this validation metric, we found that our model performed reasonably well, with an $R^2$ of 0.778 and an MAE of 0.073 for moderate food insecurity and an $R^2$ of 0.696 and a MAE of 0.044 for severe food insecurity.

For a detailed description, implementation and validation of the random forest models see the Supplemental Info.

\section{Conclusion}

@@

\section*{Supplemental Info}
\setcounter{table}{0}
\setcounter{figure}{0}
\setcounter{section}{0}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}

\section{Overview of the FIES}
We used raw microdata released by the FAO from 75 countries that vary by world region and income level (See Fig. \ref{fig:fies_countries}.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/FIES_Countries.pdf}
  \caption{Countries that had microdata on the FIES and were used in our model, shown in blue.}
  \label{fig:fies_countries}
\end{figure}

This microdata comes with the probability that an individual is over the threshold for moderate and severe food insecurity already calculated by the FAO, based on their responses to the 8 FIES questions (See List \ref{itm:fies}).  Thus, we did not perform those calculations ourselves.  Nevertheless, we give here an overview of the procedure.  For more detail see Cafiero (2018) in \textit{Measurement} \citep{Cafiero2018}.

The FIES is calculated using a Rash model, which was developed by the psychometrics literature.  The Rash model assumes that each individual and their responses to the FIES questions can be placed on a one-dimensional scale of food insecurity, and that the log odds of a respondent \textit{r} answering yes to one of the FIES questions \textit{i} is a linear function of the difference between the severity of the food insecurity experienced by \textit{r} and the severity of item {i}.

The severity of each item and each respondents level of food insecurity can be estimated with Maximum Likelihood.  After fitting the model to each country, the FAO found that the assumptions of the Rash model held up well, as evidenced by model fit diagnostics.

After separately estimating the Rash model for each country, the FAO then developed a global reference scale for the severity of each of the 8 questions, by iteratively harmonizing the severity values in each country.  Then, these global reference points were calibrated against other surveys that had asked questions similar to the FIES, such as the HFSSM and the ELCSA.

Based on the global reference scale, two thresholds are set based on questions 1 and 8: whether the responded was worried they would not have enough food to eat because of a lack of money or other resources, and whether there was a time when they went without eating for a whole day because of a lack of money or other resources.

%%% Discuss reference scale and calibration
%%% Map of countries with Training Data
\begin{enumerate}
	\item During the last 12 MONTHS, was there a time when you were worried you would not have enough food to eat because of a lack of money or other resources?
	\item Still thinking about the last 12 MONTHS, was there a time when you were unable to eat healthy and nutritious food because of a lack of money or other resources?
	\item Was there a time when you ate only a few kinds of foods because of a lack of money or other resources?
	\item Was there a time when you had to skip a meal because there was not enough money or other resources to get food?
	\item Still thinking about the last 12 MONTHS, was there a time when you ate less than you thought you should because of a lack of money or other resources?
	\item Was there a time when your household ran out of food because of a lack of money or other resources?
	\item Was there a time when you were hungry but did not eat because there was not enough money or other resources for food?
	\item During the last 12 MONTHS, was there a time when you went without eating for a whole day because of a lack of money or other resources?
  \label{itm:fies}
\end{enumerate}

\section{Predicting Covariates Into the Future}
\subsection{Annualized Rate of Change}
For many of our covariates, to extrapolate recent trends to the future, it is necessary to draw on annualized rates of change, which are then used to anticipate the rate of future change.  This method can be used for any variable that is a fraction or a rate.  We use this method to estimate future rates of stunting, wasting, and malaria mortality, as well as to model future subnational shares of GDP and population.  Thus, we give an overview of the methodology here.

First, for each subnational area, $a$, the rate of change (ROC) is calculated between each pair of adjacent years, $y$, based on a value, $p$, such that $0 < p < 1$, available for each year.

\begin{equation}
  \text{ROC}_{a,y} = \text{logit} \left( \frac{p_{a,y}}{p_{a,y-1}} \right)
  \label{eqn:a}
\end{equation}

Then, each ROC is weighted to give more weight to recent years, such that weight $W$ at year $y$ is.  For a dataset with observations for 2010 to 2017, that would be:

\begin{equation}
  W_y = \frac{y - 2010}{\sum_{y=2010}^{2017} y - 2010}
  \label{eqn:b}
\end{equation}

Then, the annualized rate of change (AROC) is calculated as:

\begin{equation}
  \text{AROC}_{a} = \text{logit} \left( \sum_{y=2010}^{2017} W_y \times \text{AROC}_{a, y} \right)
  \label{eqn:c}
\end{equation}

Finally, the projections, Proj, for each year from 2018 to 2030 are given as:

\begin{equation}
  \text{Proj}_{a,y} = \text{logit}^{-1} ( \text{logit} ( p_{a,2017} ) + AROC_{a} \times ( y - 2017 ) )
  \label{eqn:d}
\end{equation}

For cases where we are modeling shares of a whole, for example when modeling subnational shares of population or GDP, we then re-scale the shares to ensure they total to 1.

\subsection{Population}
Although population is not a covariate in our random forest model, it is an important precursor to other covariates, such as GDP per capita, and future estimates of population are also necessary to convert modeled future rates of food insecurity into total headcounts of food insecurity.  Thus, in addition to our other covariates, we also modeled population into the future.

We do this by combining historic subnational data from the Subnational Development Database \citep{Smits2019} with national level population estimates for the 21st century \citep{KC2017}.  To account for within-country trends and changes in population distribution, we use the AROC method outlined in equations \ref{eqn:a} - \ref{eqn:d} to project each subnational areas share of national population totals. We then disaggregate the national totals given by KC et al. to estimate future subnational populations.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/population.png}
  \caption{Population}
\end{figure}

\subsection{Urban Percentage}

We drew our historic and future estimates of urbanization entirely from Jones and O'Neill, published in \textit{Environmental Research Letters} \citep{Jones2016}, who modeled spatially explicit scenarios of urbanization consistent with the Shared Socioeconomic Pathways.  As with our other covariates, we used estimates for SSP2, the middle-of-the-road scenario.  Because definitions of ``urban" and ``rural" vary significantly across datasets, using one dataset that is entirely consistent within itself was an important consideration.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/urban_perc.png}
  \caption{Percentage of people living in urban areas.}
\end{figure}

\subsection{Wasting}
To estimate the prevalence of wasting for each administrative area in our dataset, we used data from the Local Burden of Disease project published in \textit{Nature} \citep{Local2020}.  For the years 2010 to 2017, we simply took the mean rate of wasting in each administrative area from the dataset.  Higher-income countries that were not included in the dataset were modeled as having a rate of 0 wasting.  We then modeled wasting for the years 2018-2030 using the AROC method, which the Local Burden of Disease group similarly used to estimate wasting for the year 2025 \citep{Local2020}.

To account for the effects of the novel coronavirus on global rates of wasting, we assumed that long-term trends and rates of wasting would hold steady, but that they increase globally by 14.3\% in 2020, based on estimates published in \textit{The Lancet} \citep{headey2020impacts}.  We model this impact as uniform across all countries where wasting occurs.  Then we model the rate of wasting in each administrative area in 2021 as being the mean of the rate in 2020 that accounts for the shock and the previously modeled rate for 2022.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/wasting.png}
  \caption{Prevalence of wasting.}
\end{figure}

\subsection{Stunting}

We model stunting using the same methodology we use to summarize and model wasting, because the Local Burden of Disease dataset that we draw on for wasting also includes stunting.  We also included the 14.3\% increase in prevalence for the year 2020 that was estimated for wasting \citep{headey2020impacts}, even though stunting is a long-term consequence of malnutrition and will likely not be as readily observable in a population as wasting would be.  However, we use stunting in our model not as a cause of food insecurity, but rather as a proxy for chronic conditions of hunger, which are exacerbated by the coronavirus pandemic.  Thus, estimating an impact of the coronavirus shock on stunting is appropriate for our model.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/stunting.png}
  \caption{Prevalence of stunting.}
\end{figure}

\subsection{GDP Per Capita}
We estimate global subnational GDP per capita using historic subnational estimates of Gross National Income (GNI) from the Subnational Human Development Database \cite{Smits2019} and forecasts of national GDP within the SSP framework from Dellink et al. \cite{Dellink2017}.  We first use data from the World Bank to estimate the mean country-level ratio of GNI to GDP and adjust the subnational estimates of GNI according to that country-specific ratio.  We then calculate each subnational area's share of country GDP, and then use the AROC method to estimate each subnational area's share of total country GDP.  Based on these trajectories in each subnational areas share of total country GDP, we estimate what each subnational area's share of GDP will be for each year from 2019 to 2030.  We divide that share of GDP by the population of the administrative area, which we had previously estimated.  Thus, national GDP totals are consistent with the projections of Dellink et al., but are disaggregated according to the historic patters of GDP share in the Subnational Human Development Database.

Finally, we used estimates of country-level changes in GDP growth for the years 2020 and 2021 to account for the impact of the coronavirus \citep{prospects2020}.  We then model GDP as resuming its previous growth trajectory based on previously-estimated growth rates from 2022-2030, but growing from an estimate for 2021 that includes the coronavirus shock.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/gdp_percap.png}
  \caption{GDP per capita.}
\end{figure}

\subsection{Mean Years of Schooling}
We estimate global subnational mean years of schooling using historic subnational estimates from the Subnational Human Development Database \cite{Smits2019} and forecasts of national mean years of schooling from KC et al. \cite{KC2017}.  Using the historic subnational data, we estimate each subnational administrative area's difference in years of schooling from the national mean, and use this value to disaggregate the future projections to a subnational level.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/school_mean.png}
  \caption{Mean years of schooling.}
\end{figure}

\subsection{Gini Coefficient}
We use estimates of national income inequality from Rao et al. \citep{Rao2019a}.  These estimates include all years in our analysis, and are not disaggregated to a subnational level.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/gini.png}
  \caption{Gini Coefficient}
\end{figure}

\subsection{Poverty Headcount Index}
We used estimates of the Poverty Headcount Index for the recent past and future from the World Poverty Clock by the World Data Lab \citep{Cuaresma2018}, updated to include the impact of the coronavirus. In our model we include national level poverty rates at a threshold of \$1.90 which is defined as \textit{extreme} poverty. By using the concept of parameterized Lorenz curves in combination with mean income/consumption and population projections future poverty rates are calculated. For more details on the methodology and information on global and country level poverty check the website \url{worldpoverty.io}.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/hci.png}
  \caption{Poverty Headcount Index}
\end{figure}

\subsection{Water Scarcity Share}
We used the water scarcity index (WSI) developed by \citep{greve2018global}. The index represents the ratio of the decadal averages in water demand to water supply and it assesses changes in water scarcity on a 0.5 by 0.5 degree global grid.  Based on this index, we estimate the share of people living in areas with less than 1000m\textsuperscript{3} of water available per capita per year. The WSI has been used already to build the water scarcity clock. For more information visit \href{https://worldwater.io/}{worldwater.io}.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/ws_share.png}
  \caption{Water Scarcity Share}
\end{figure}

\subsection{Mean Annual Precipitation}
For mean annual precipitation, we use data for the years 2010 to 2019 from the TerraClimate dataset \citep{abatzoglou2018terraclimate}, summarized to each subnational area.  For data for the years 2020-2030, we use models from the Inter-Sectoral Model Intercomparison Project \citep{warszawski2014inter}.  Specifically, we use an ensemble of bias-corrected projections under representative concentration pathway (RCP) 6.0, taking the mean of the ensemble.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/precip.png}
  \caption{Mean annual precipitation}
\end{figure}

\subsection{Mean Temperature}
For mean temperature, we used the same datasets that we'd used for precipitation, as each of those datasets included both temperature in addition to precipitation.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/tave.png}
  \caption{Mean temperature (Celsius)}
\end{figure}

\subsection{Topographic Ruggedness}
As a measure of accessibility, we include the mean topographic ruggedness of each subnational area.  We first used a gridded dataset of elevation from the USGS \cite{USGS1996} and calculated the index at each grid cell using the methodology from Riley et al. \citep{Riley1999}.  We then summarized the values of all the grid cells within each administrative area.  Because this variable does not change over time, it is constant across all years in the model.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/ruggedness.png}
  \caption{Topographic ruggedness}
\end{figure}

\subsection{Malaria (\textit{P. falciparum}) Mortality Rate}
We used data from \textit{The Lancet} to estimate the mortality (deaths per 100,000 population per year) due to Malaria \citep{Weiss2019}, taking the average pixel values within each subnational administrative area.  We then estimated future malaria mortality based on AROC method.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/mal_falciparum.png}
  \caption{Rate of mortality due to \textit{P. falciparum} Malaria}
\end{figure}

\section{Description, Implementation and Validation of the RF regression model}

The RF regression is a machine learning method based on the creation of a large number of decision trees. By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to single decision trees \citep{friedman2001elements}.  For each of the decision trees in a random forest model, observations are selected at random with replacement, a method known as bootstrap aggregating, or bagging, and features are area also selected at random.

For implementation we use the R-package \texttt{randomForestSRC} by Ishwaran and Kogalur \citep{ishwaran2019randomforestsrc}. Specifically, we apply the functions \texttt{tune.rfsrc()} to find the optimal combination of hyperparamters and the function \texttt{rfsrc()} to perform the RF regression models. After predicting rates of moderate-to-severe and severe food insecurity we plot the modeled vs. observed rates and calculated the MSE and ${R}^2$ (See Fig. \ref{fig:rf_in-sample} and Fig. \ref{fig:rf_out-sample}) for both models. Additionally, we use the function \texttt{vimp()} to gain insights into the development of the error rates with increasing numbers of trees (See Fig. \ref{fig:rf_error}) and the importance of individual variables (See Fig. \ref{fig:rf_vimp}).

%STEP-BY-STEP: LOGIT TRANSFORM, TUNE, MODEL, PREDICT, VALIDATE
Because the outcome variables in our models, the rate of moderate or severe food insecurity, are a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.  Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.

To ensure that predicted rates are bounded by 0 and 1 we conduct a logit transformation of the dependent variable and inverse logit transformation of the predictions. 
%MAYBE STATE THIS SOMEWHERE ELSE, NOT SURE WHERE

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/error_rf.png}
  \caption{Error Rate of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_error}
\end{figure}

%HERE A FEW SENTENCES ON THE NUMBER OF TREES WITH LINK TO NEXT PARAGRAPH

In addition to the parameter that describes the number of trees to be created, two other important hyperparameters must be tuned. In the function \texttt{tune.rfsrc()} these parameters are called \texttt{mtry} and \texttt{nodesize}. \texttt{mtry} describes the number of variables randomly selected as candidates for splitting a node and \texttt{nodesize} describes the average number of observations in a leaf node. The optimal pair of hyperparamters is found by trying different combinations and choosing the one with the smallest out-of-bag (OOB) error.

%EXPLAIN THE OOB ERROR HERE?

The function \texttt{tune.rfsrc()} does exactly that and returns optimal values for \texttt{mtry} and \texttt{nodesize}. We applied this function on both, the moderate-to-severe and the severe model. For the moderate-to-severe model the optimal combination is \texttt{mtry} = 12 and \texttt{nodesize} = 1 and for the severe model we found \texttt{mtry} = 10 and \texttt{nodesize} = 2 to be optimal.

%STATE FUNCTION ARGUMENTS HERE?

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/in-sample_rf.png}
  \caption{In-Sample Fit of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_in-sample}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/out-sample_rf.png}
  \caption{Out-Of-Sample Fit of the RF regression model with 75\% training set and 25\% test set. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_out-sample}
\end{figure}

%HERE EVERYTHING ON THE MODEL, NOT SURE HOW DETAILED
%For modeling the rates of moderate-to-severe and severe food insecurity we use the function \texttt{rfsrc()}. We specify

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/VIMP.pdf}
  \caption{Variable Importance of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_vimp}
\end{figure}

%https://kogalur.github.io/randomForestSRC/theory.html#section4.2
Lastly we asses the importance of individual variables. This is based on the Breiman-Culter permutation variable importance information \citep{breiman2001random}.  This involved comparing the prediction error on the OOB data to the prediction error of OOB cases where the x-variable $x$ is randomly permutated.  The importance of an individual variable is thus the mean difference between the perturbed and unperturbed error rate, across all trees. 
%MORE INFORMATION HERE
\end{document}

>>>>

When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.  We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 3-7 percentage points on average when applied to countries outside of the training set.

@@

\subsection{Modeling}
We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.  We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.  A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.  By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}. Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.  We create our models using the \texttt{randomForestSRC} package in R \citep{ishwaran2019randomforestsrc}.

To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.  These include the average number of observations in a leaf node, the number of variables randomly selected as candidates for splitting a node, and the maximum depth of each tree.  We use the combination of hyperparameters that perform best under ten-fold cross validation based on $R^2$.

When performing cross validation, it is typical to select observations at random.  However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.  Thus, for each iteration of cross validation, we leave out data from 10\% of the countries in our dataset, rather than 10\% of the total observations.  Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.  By this validation metric, we find that our models perform well, with an $R^2$ of 0.799 and mean average error (MAE) of 0.068 for moderate food insecurity and an $R^2$ of 0.772 and a MAE of 0.038 for severe food insecurity.  After fitting models on the entire dataset, we observe an $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.944 and a MAE of 0.019 for severe food insecurity.

After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity. We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable, breiman2001random}.  These metric can provide some interpretability to model predictions.

The Supplementary Materials provide a detailed description of the implementation and validation of the random forest models.

@@
